A unified pseudo-$C_\ell$ framework

### 2018-08-30 doc/Paper/main.tex
another small typo
addressed most of Reviewer 3's comments

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information, their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ and $\ell_{\rm max}^4$ respectively with the maximum multipole $\ell_{\rm max}$. For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $E/B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, while a number of power spectrum codes solving some of these problems have been put forward, no public, validated and well-maintained code exists supporting all of these capabilities consistently. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

@@

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \va^\dagger \tilde{\sf C}^{-1}\,{\sf P}_i\,\tilde{\sf C}^{-1} \va -
          \rm{Tr}\left[{\sf P}_i\,{\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix,the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}\label{eq:bpws_qmv}
          \tilde{\sf C} = \sum_i {\sf P}_i\,B_i + {\sf N},
        \end{equation}
        and the last term in Eq. \ref{eq:bpws_qmv} correspond to bias of convolved estimates due to the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices ${\sf P}_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

      \subsubsection{Mode coupling}\label{sssec:maths.pcl101.mcoup}
        Let $v(\nv)$ be a sky mask or weights map for ${\bf a}$, and let us define ${\bf a}^v\equiv v(\nv){\bf a}(\nv)$. As discussed above we start by considering a na\"ive estimator for ${\sf C}^{ab}_\ell$ from the masked fields:
        \begin{equation}\label{eq:pcldef}
          {\rm PCL}_\ell\left({\bf a}^v,{\bf b}^w\right)\equiv\frac{1}{2\ell+1}\sum_{m=-\ell}^\ell{\bf a}^v_{\bf l}{\bf b}^{w\dag}_{\bf l}.
        \end{equation}
        The incomplete sky coverage couples different $\ell$ modes, and makes this estimator biased. The pseudo-$C_\ell$ method is based on computing an analytical prediction for this bias and correcting for it. This is straightforward to do using the results in Appendix \ref{app:shts}. The harmonic coefficients of a masked field are given by
        \begin{equation}
          {\bf a}^v_{\bf l}=\sum_{{\bf l}_1{\bf l}_2}={\sf D}^{s_a}_{{\bf l}{\bf l}_1{\bf l}_2}{\bf a}_{{\bf l}_1}v_{{\bf l}_2},
        \end{equation}
        where ${\sf D}$ is defined in Eq. \ref{eq:dmat}. Using the statistical isotropy of the unmasked field (Eq. \ref{eq:cldef}) and the orthogonality relation Eq. \ref{eq:3jorth}, we obtain:
        \begin{equation}\label{eq:modecoup}
          {\rm vec}\left[\left\langle{\rm PCL}_\ell\left({\bf a}^v_{\bf l},{\bf b}^w_{\bf l}\right)\right\rangle\right]=\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\cdot{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right],
        \end{equation}
        where we have defined the vectorization operator \citep{2008PhRvD..77j3013H}
        \begin{align}
          &{\rm vec}\left[\left(\begin{array}{cc}
                          C^{EE} & C^{EB}\\
                          C^{BE} & C^{BB}
                         \end{array}\right)\right]\equiv
          \left(
          \begin{array}{c}
            C^{EE} \\
            C^{EB} \\
            C^{BE} \\
            C^{BB} \\
          \end{array}
          \right),\\
          &{\rm vec}\left[\left(\begin{array}{c}
                                 C^{TE}\\C^{TB}
                                \end{array}\right)\right]\equiv
            \left(\begin{array}{c}
                    C^{TE}\\C^{TB}
            \end{array}\right),
            \hspace{6pt}
            {\rm vec}\left[C^{TT}\right]={\rm C}^{TT},
        \end{align}
        and $T$ here stands for any spin-$0$ field.

        The mode-coupling matrix ${\sf M}_{\ell\ell'}$ in Eq. \ref{eq:modecoup} can be computed in terms of the harmonic coefficients of the two masks and, for spin-0 and spin-2 fields, is given by:
        \begin{align}\label{eq:m00}
          &{\sf M}^{00}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{0}{0}{0}^2\,\mI,\\\label{eq:m02}
          &{\sf M}^{02}_{\ell\ell'}=M^{0+}_{\ell\ell'}\,\mI,\\\nonumber
          &M^{0+}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{0}{0}{0}\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0},\\\label{eq:m22}
          &{\sf M}^{22}=\left(
                        \begin{array}{cccc}
                          M^{+}_{\ell\ell'} & 0 & 0 & M^{-}_{\ell\ell'} \\
                          0 & M^{+}_{\ell\ell'} & -M^{-}_{\ell\ell'} & 0 \\
                          0 & -M^{-}_{\ell\ell'} & M^{+}_{\ell\ell'} & 0 \\
                          M^{-}_{\ell\ell'} & 0 & 0 & M^{+}_{\ell\ell'}
                        \end{array}\right),\\\nonumber
          &M^{\pm}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}^2\frac{1\pm(-1)^{\ell+\ell'+\ell''}}{2},
        \end{align}
        where $\mI$ is the unit matrix, and $P^{vw}_\ell\equiv(2\ell+1)\,{\rm PCL}_\ell(v,w)$.

@@

    \subsection{Contaminant deprojection}\label{ssec:maths.deproj}
      In a typical data-analysis case, we deal with contaminants that pollute the measured signal. Usually, we know something about these contaminants. As an example, in case of CMB analysis, we might have have a map of a given foreground at a different frequency which can be marginalized out. But even when we do not, we often have a tracer map of a given effect, for example, the map of dust might tell us something about systematic errors related to imperfect correction to dust reddening, or a map of co-added PSF fluctuations could be used to remove the possible correlations between measured ellipticities. Even if the dust map is non-linearly and perhaps stochastically related to the amount of reddening, as long as relation is local in space, we get a linear contamination on large scales, very much like complex astrophysics of galaxy formation nevertheless reduces to linear biasing on very large scales (see Appendix \ref{app:lincont}). Thus, we generically deal with a linear model of the kind:
      \begin{equation}\label{eq:aobs}
        {\bf a}^{\rm obs}(\nv)={\bf a}^v+\sum_{i=1}^{N_a}\alpha_i{\bf f}^i(\nv),
      \end{equation}
      where the observed map ${\bf a}^{\rm obs}$ is composed of a masked signal map that receives contributions from $N_a$ known contaminants ${\bf f}^i$ and where the contamination coefficients $\alpha_i$ are unknown (note that we have implicitly applied the same mask to the contaminant templates ${\bf f}^i$). The deprojection of these contaminants from the data can be done and propagated into the estimated power spectrum in a natural way in the case of quadratic estimators \citep{1992ApJ...398..169R,2004PhRvD..69l3003S,2016MNRAS.456.2095E}, and the method can be extended to the pseudo-$C_\ell$ algorithm \citep{2017MNRAS.465.1847E}. The process is illustrated in Figure \ref{fig:lss_cont}. A best-fit value for the coefficients $\alpha_i$ can be found assuming \emph{equal weights} across the map\footnote{The assumption of uncorrelated weights is the key difference between pseudo-$C_\ell$ and optimal quadratic estimators \citep{2013MNRAS.435.1857L}. Note that, since the templates are already multiplied by the mask, the weights are not equal across the map, but correspond to the mask values in general.}
      \begin{align}\label{eq:deproj_coeff}
        &\hat{\alpha}_i=F_{ij}\int d\nv\,{\bf f}^{j\dag}(\nv)\,{\bf a}^{\rm obs}(\nv),\\\nonumber
        &({\sf F}^{-1})_{ij}\equiv\int d\nv\, {\bf f}^{i\dag}(\nv)\,{\bf f}^j(\nv),
      \end{align}
      where there is an implicit summation over the repeated index $j$ in Eq. \ref{eq:deproj_coeff} (which we will omit from now on). The clean map is then given by
      \begin{align}
        {\bf a}^c(\nv)
        &\equiv{\bf a}^{\rm obs}(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^{\rm obs}(\nv')
      \end{align}
      which, when substituting Eq. \ref{eq:aobs}, yields
      \begin{align}\label{eq:clean}
        {\bf a}^c(\nv)&={\bf a}^v(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^v(\nv').
      \end{align}      
      The harmonic coefficients of ${\bf a}^c$ are therefore given by
      \begin{equation}\label{eq:clean_f}
        {\bf a}^c_{\bf l}={\bf a}^v_{\bf l}-{\bf f}^i_{\bf l}\,F_{ij}\sum_{{\bf l}'}{\bf f}^{j\dag}_{{\bf l}'}{\bf a}^v_{{\bf l}'}
      \end{equation}

@@

      Two further points must be addressed before moving on. First, the computation of the deprojection bias requires an estimate of the true power spectrum ${\sf C}^{ab}_\ell$ (see Eqs. \ref{eq:dp.g.aux} and \ref{eq:dp.f.aux}). In practice, this can be done through an iterative approach, starting from an initial guess of ${\sf C}^{ab}_\ell$ and replacing it by the estimated power spectrum in the previous iteration. A sufficiently good guess will likely be enough to remove the bias in a single iteration.  Alternatively, the ensemble average of Eq. \ref{eq:dpbias0} can be seen as a convolution of ${\sf C}^{ab}_\ell$, which could be treated by computing the convolution matrix as a modified version of the standard mode-coupling matrix that accounts for mode deprojection. Estimating this convolution matrix in this case becomes significantly more computationally intensive, and therefore an iterative approach is preferred.

      Secondly, estimating the deprojection coefficients (Eq. \ref{eq:deproj_coeff}) involves inverting the correlation matrix of all the contaminant templates. This will not be in general possible (e.g. if a set of templates are linearly related). The problem can however be solved self-consistently by using the Moore-Penrose pseudo-inverse. This is trivial for a symmetric and positive-definite matrix such as ${\sf F}$, and corresponds to setting the inverse of all of its zero eigenvalues (or all those below a given threshold) to zero.

    \subsection{\texorpdfstring{$E/B$}{E/B} purification}\label{ssec:maths.pureb}
      As described in \cite{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2010PhRvD..82b3001Z,2010A&A...519A.104K,2011PhRvD..83h3003B}, $B$-mode purification refers to the map-level removal of the contamination from $E$-modes in a the $B$-mode component of a given map caused by an incomplete sky coverage and vice-versa\footnote{Note that this is a different effect from the leakage between different polarized and unpolarized channels in CMB observations due to instrumental effects (asymmetric beams, pair differencing, polarization angle calibration etc.)}. The procedure is particularly useful in situations in which the $E$-mode component of the signal is significantly larger than the $B$ modes, as is the case in CMB. In this case, removing the leakage at the power spectrum level (i.e. the standard pseudo-$C_\ell$ approach) produces a sub-optimal estimator in which the variance in the $B$-mode power spectrum is dominated by the variance of the leaked $E$-modes. This section presents the purification method and the associated modifications to the pseudo-$C_\ell$ algorithm as described in the references above. In this section we describe the algorithm in terms of purifying $B$-modes from $E$-mode contamination, but we note that the reverse, i.e. purifying $E$-modes from $B$-mode contamination is conceptually exactly the same. Our discussion is also specific for the spin-2 fields.

@@

      Since ${\bf D}^{E\dag}_s{\bf D}^B_s=0$, one can always generate a $B$ mode by applying ${\bf D}^{B}_s$ to a scalar field. It is then possible to show that $E$ and $B$ modes thus defined are orthogonal in the full sky:
      \begin{equation}
        ({\bf D}^E_s\varphi,{\bf D}^B_s\psi)=\int d\nv\,({\bf D}^E_s\varphi)^\dag{\bf D}^B_s\psi=0
      \end{equation}
      This can be done by integrating by parts and noting that the celestial sphere has no boundaries. On a cut sky, however, and for $s=2$, this is only true if the fields satisfy Neumann and Dirichlet boundary conditions simultaneously (i.e. vanishing value and first derivative on the boundary of the cut sky region).

      Let $w(\nv)$ be a sky window function defining the sky region to be analyzed (and the weight to be applied in each pixel). The standard pseudo $B$-mode of a spin-2 field ${\bf P}$ is then given by
      \begin{align}\nonumber
        \tilde{B}_{\bf l}&\equiv\int d\nv\,w(\nv)\left(_s{\bf Y}^{B}_{\bf l}(\nv)\right)^\dag{\bf P}\\
                         &=\int d\nv\,w(\nv)({\bf D}^B_sY_{\bf l})^\dag{\bf P}(\nv),
      \end{align}
      Since ${\bf D}^B_sY_{\ell m}$ is a $B$-mode, in the absence of $w$ this expression would correspond to a projection that filters out all the $E$-modes from ${\bf P}$. However, $w(\nv){\bf D}^B_sY_{\ell m}$ is not a $B$-mode, and therefore $\tilde{B}_{\ell m}$ receives contributions from ambiguous $E$ modes (which then propagate into the variance of the pseudo-$C_\ell$ estimator of the power spectrum). The idea behind $B$-mode purification is to move $w$ to the right of ${\bf D}^B_s$, defining the pure $B$ component:
      \begin{equation}
        B^p_{\bf l}=\int d\nv\left({\bf D}^B_2(w\,Y_{\bf l})\right)^\dag\,{\bf P}(\nv).
      \end{equation}
      Since ${\bf D}^B_2(wY_{\bf l})$ is a $B$-mode quantity, $B^p_{\bf l}$ should receive contributions only from $B$-modes.

@@

  \section{Code performance and validation} \label{sec:validation}
    The core of our code-validation suite is based on two science examples. These examples are not meant to be realistic mocks of actual data, however, they do contain the dynamic range and noise-levels of a typical experimental set-up that we expect in the next decade, while providing us with an exact, known true power spectrum to compare against. This section presents these two examples and the results of this validation in terms of accuracy and computational performance. Results will be presented for both curved-sky and flat-sky realizations.

@@

          \item {\bf Contaminants:} we contaminate the signal maps described above with several different types of residuals. In all cases we add contaminants linearly, to avoid deviating from the model described in Section \ref{ssec:maths.deproj}. Maps of these contaminants are shown in the 4 bottom panels of Figure~\ref{fig:lss_mocks_sph} and Figure~\ref{fig:lss_mocks_flat}, and their impact on the angular power spectrum is presented in Fig. \ref{fig:lss_cont_cls} for the curved-sky case.

>>>>

    While maximum-likelihood~\citep{1998PhRvD..57.2117B,2003PhRvD..67b3001W} or minimum-variance and quadratic estimators~\citep{1997PhRvD..55.5895T,2000MNRAS.317L..23H,2001PhRvD..64f3001T} can recover the power spectrum with virtually no loss of information (see also the approaches of e.g. \cite{2004ApJS..155..227E,2008MNRAS.389.1284T,2016MNRAS.455.4452A}, based on Bayesian sampling methods), their computational implementation is prohibitively expensive for high resolution data, since these methods scale as $\ell_{\rm max}^6$ with the maximum multipole $\ell_{\rm max}$ (or $\ell_{\rm max}^4$ for optimal algorithms such as that of \cite{2003PhRvD..67b3001W}). For this reason, pseudo-$C_\ell$ algorithms \citep{1973ApJ...185..413P,2001PhRvD..64h3003W,2001ApJ...561L..11S,2001PhRvD..64h3003W,2002ApJ...567....2H,2002MNRAS.336.1304H,2004MNRAS.350..914C} have become a popular alternative that preserves the $\ell_{\rm max}^3$ scaling of spherical harmonic transforms (and faster in the flat-sky approximation). This method can be used on fields with arbitrary spin \citep{2003ApJS..148..161K} in both curved and flat skies, has been extended to deal with contaminant deprojection in scalar fields \citep{2017MNRAS.465.1847E}, and can be optimized to measure the CMB $E/B$-mode power spectrum \citep{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2009PhRvD..79l3515G,2011PhRvD..83h3003B}. Unfortunately, in spite of the advantages of this method, while a number of power spectrum codes solving some of these problems have been put forward, no public, validated and well-maintained code exists supporting all of these capabilities consistently. This paper presents a public software package, \nmt, that unifies and extends all of these features providing a general framework to estimate pseudo-$C_\ell$ angular power spectra in a wide variety of scenarios.

@@

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \frac{1}{2}\va^\dagger \tilde{\sf C}^{-1}\,{\sf P}_i\,\tilde{\sf C}^{-1} \va -
          \frac{1}{2}\rm{Tr}\left[\tilde{\sf C}^{-1}{\sf P}_i\tilde{\sf C}^{-1}\,{\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix, the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}\label{eq:bpws_qmv}
          \tilde{\sf C} = \sum_i {\sf P}_i\,B_i + {\sf N},
        \end{equation}
        and the last term in Eq. \ref{eq:bpws_qmv} correspond to bias of convolved estimates due to the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices ${\sf P}_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

      \subsubsection{Mode coupling}\label{sssec:maths.pcl101.mcoup}
        Let $v(\nv)$ be a sky mask or weights map for ${\bf a}$, and let us define ${\bf a}^v\equiv v(\nv){\bf a}(\nv)$. As discussed above we start by considering a na\"ive estimator for ${\sf C}^{ab}_\ell$ from the masked fields:
        \begin{equation}\label{eq:pcldef}
          {\rm PCL}_\ell\left({\bf a}^v,{\bf b}^w\right)\equiv\frac{1}{2\ell+1}\sum_{m=-\ell}^\ell{\bf a}^v_{\bf l}{\bf b}^{w\dag}_{\bf l}.
        \end{equation}
        The incomplete sky coverage couples different $\ell$ modes, and makes this estimator biased. The pseudo-$C_\ell$ method is based on computing an analytical prediction for this bias and correcting for it. This is straightforward to do using the results in Appendix \ref{app:shts}. The harmonic coefficients of a masked field are given by
        \begin{equation}
          {\bf a}^v_{\bf l}=\sum_{{\bf l}_1{\bf l}_2}{\sf D}^{s_a}_{{\bf l}{\bf l}_1{\bf l}_2}{\bf a}_{{\bf l}_1}v_{{\bf l}_2},
        \end{equation}
        where ${\sf D}$ is defined in Eq. \ref{eq:dmat}. Using the statistical isotropy of the unmasked field (Eq. \ref{eq:cldef}) and the orthogonality relation Eq. \ref{eq:3jorth}, we obtain:
        \begin{equation}\label{eq:modecoup}
          {\rm vec}\left[\left\langle{\rm PCL}_\ell\left({\bf a}^v_{\bf l},{\bf b}^w_{\bf l}\right)\right\rangle\right]=\sum_{\ell'}{\sf M}^{s_as_b}_{\ell\ell'}\cdot{\rm vec}\left[{\sf C}^{ab}_{\ell'}\right],
        \end{equation}
        where we have defined the vectorization operator \citep{2008PhRvD..77j3013H}
        \begin{align}
          &{\rm vec}\left[\left(\begin{array}{cc}
                          C^{EE} & C^{EB}\\
                          C^{BE} & C^{BB}
                         \end{array}\right)\right]\equiv
          \left(
          \begin{array}{c}
            C^{EE} \\
            C^{EB} \\
            C^{BE} \\
            C^{BB} \\
          \end{array}
          \right),\\
          &{\rm vec}\left[\left(\begin{array}{c}
                                 C^{TE}\\C^{TB}
                                \end{array}\right)\right]\equiv
            \left(\begin{array}{c}
                    C^{TE}\\C^{TB}
            \end{array}\right),
            \hspace{6pt}
            {\rm vec}\left[C^{TT}\right]={\rm C}^{TT},
        \end{align}
        and $T$ here stands for any spin-$0$ field.

        The mode-coupling matrix ${\sf M}_{\ell\ell'}$ in Eq. \ref{eq:modecoup} can be computed in terms of the harmonic coefficients of the two masks and, for spin-0 and spin-2 fields, is given by \citep{2002ApJ...567....2H,2003ApJS..148..161K}:
        \begin{align}\label{eq:m00}
          &{\sf M}^{00}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{0}{0}{0}^2\,\mI,\\\label{eq:m02}
          &{\sf M}^{02}_{\ell\ell'}=M^{0+}_{\ell\ell'}\,\mI,\\\nonumber
          &M^{0+}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{0}{0}{0}\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0},\\\label{eq:m22}
          &{\sf M}^{22}=\left(
                        \begin{array}{cccc}
                          M^{+}_{\ell\ell'} & 0 & 0 & M^{-}_{\ell\ell'} \\
                          0 & M^{+}_{\ell\ell'} & -M^{-}_{\ell\ell'} & 0 \\
                          0 & -M^{-}_{\ell\ell'} & M^{+}_{\ell\ell'} & 0 \\
                          M^{-}_{\ell\ell'} & 0 & 0 & M^{+}_{\ell\ell'}
                        \end{array}\right),\\\nonumber
          &M^{\pm}_{\ell\ell'}=\frac{2\ell'+1}{4\pi}\sum_{\ell''}P^{vw}_{\ell''}\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}^2\frac{1\pm(-1)^{\ell+\ell'+\ell''}}{2},
        \end{align}
        where $\mI$ is the unit matrix, and $P^{vw}_\ell\equiv(2\ell+1)\,{\rm PCL}_\ell(v,w)$.

@@

    \subsection{Contaminant deprojection}\label{ssec:maths.deproj}
      In a typical data-analysis case, we deal with contaminants that pollute the measured signal. Usually, we know something about these contaminants. As an example, in case of CMB analysis, we might have have a map of a given foreground at a different frequency which can be marginalized out. But even when we do not, we often have a tracer map of a given effect, for example, the map of dust might tell us something about systematic errors related to imperfect correction to dust reddening, or a map of co-added PSF fluctuations could be used to remove the possible correlations between measured ellipticities. Even if the dust map is non-linearly and perhaps stochastically related to the amount of reddening, as long as relation is local in space, we get a linear contamination on large scales, very much like complex astrophysics of galaxy formation nevertheless reduces to linear biasing on very large scales (see Appendix \ref{app:lincont}). Thus, we generically deal with a linear model of the kind:
      \begin{equation}\label{eq:aobs}
        {\bf a}^{\rm obs}(\nv)={\bf a}^v+\sum_{i=1}^{N_a}\alpha_i{\bf f}^i(\nv),
      \end{equation}
      where the observed map ${\bf a}^{\rm obs}$ is composed of a masked signal map that receives contributions from $N_a$ known contaminants ${\bf f}^i$ and where the contamination coefficients $\alpha_i$ are unknown (note that we have implicitly applied the same mask to the contaminant templates ${\bf f}^i$). The deprojection of these contaminants from the data can be done and propagated into the estimated power spectrum in a natural way in the case of quadratic estimators \citep{1992ApJ...398..169R,2004PhRvD..69l3003S,2016MNRAS.456.2095E}, and the method can be extended to the pseudo-$C_\ell$ algorithm \citep{2017MNRAS.465.1847E}. The process is illustrated in Figure \ref{fig:lss_cont}. A best-fit value for the coefficients $\alpha_i$ can be found assuming \emph{uncorrelated weights} across the map\footnote{The assumption of uncorrelated weights is the key difference between pseudo-$C_\ell$ and optimal quadratic estimators \citep{2013MNRAS.435.1857L}. Note that, since the templates are already multiplied by the mask, the weights are not equal across the map, but correspond to the mask values in general.}
      \begin{align}\label{eq:deproj_coeff}
        &\hat{\alpha}_i=F_{ij}\int d\nv\,{\bf f}^{j\dag}(\nv)\,{\bf a}^{\rm obs}(\nv),\\\nonumber
        &({\sf F}^{-1})_{ij}\equiv\int d\nv\, {\bf f}^{i\dag}(\nv)\,{\bf f}^j(\nv),
      \end{align}
      where there is an implicit summation over the repeated index $j$ in Eq. \ref{eq:deproj_coeff} (which we will omit from now on). The clean map is then given by
      \begin{align}
        {\bf a}^c(\nv)
        &\equiv{\bf a}^{\rm obs}(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^{\rm obs}(\nv')
      \end{align}
      which, when substituting Eq. \ref{eq:aobs}, yields
      \begin{align}\label{eq:clean}
        {\bf a}^c(\nv)&={\bf a}^v(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^v(\nv').
      \end{align}      
      The harmonic coefficients of ${\bf a}^c$ are therefore given by
      \begin{equation}\label{eq:clean_f}
        {\bf a}^c_{\bf l}={\bf a}^v_{\bf l}-{\bf f}^i_{\bf l}\,F_{ij}\sum_{{\bf l}'}{\bf f}^{j\dag}_{{\bf l}'}{\bf a}^v_{{\bf l}'}
      \end{equation}

@@

      Two further points must be addressed before moving on. First, the computation of the deprojection bias requires an estimate of the true power spectrum ${\sf C}^{ab}_\ell$ (see Eqs. \ref{eq:dp.g.aux} and \ref{eq:dp.f.aux}). In practice, this can be done through an iterative approach, starting from an initial guess of ${\sf C}^{ab}_\ell$ and replacing it by the estimated power spectrum in the previous iteration. Alternatively, the ensemble average of Eq. \ref{eq:dpbias0} can be seen as a convolution of ${\sf C}^{ab}_\ell$, which could be treated by computing the convolution matrix as a modified version of the standard mode-coupling matrix that accounts for mode deprojection. Estimating this convolution matrix in this case becomes significantly more computationally intensive, and therefore an iterative approach is preferred.

      Secondly, estimating the deprojection coefficients (Eq. \ref{eq:deproj_coeff}) involves inverting the correlation matrix of all the contaminant templates. This will not be in general possible (e.g. if a set of templates are linearly related). The problem can however be solved self-consistently by using the Moore-Penrose pseudo-inverse \citep{2017MNRAS.465.1847E}. This is trivial for a symmetric and positive-definite matrix such as ${\sf F}$, and corresponds to setting the inverse of all of its zero eigenvalues (or all those below a given threshold) to zero.

    \subsection{\texorpdfstring{$E/B$}{E/B} purification}\label{ssec:maths.pureb}
      As described in \cite{2002PhRvD..65b3505L,2003PhRvD..67b3501B,2006PhRvD..74h3002S,2010PhRvD..82b3001Z,2010A&A...519A.104K,2011PhRvD..83h3003B}, $B$-mode purification refers to the map-level removal of the contamination from $E$-modes in a the $B$-mode component of a given map caused by an incomplete sky coverage and vice-versa\footnote{Note that this is a different effect from the leakage between different polarized and unpolarized channels in CMB observations due to instrumental effects (asymmetric beams, pair differencing, polarization angle calibration etc.)}. The procedure is particularly useful in situations in which the $E$-mode component of the signal is significantly larger than the $B$ modes, as is the case in CMB. In this case, removing the leakage at the power spectrum level (i.e. the standard pseudo-$C_\ell$ approach) produces a sub-optimal estimator in which the variance in the $B$-mode power spectrum is dominated by the variance of the leaked $E$-modes. This section presents the purification method and the associated modifications to the pseudo-$C_\ell$ algorithm as described in the references above. In this section we describe the algorithm in terms of purifying $B$-modes from $E$-mode contamination, but we note that the reverse, i.e. purifying $E$-modes from $B$-mode contamination is conceptually exactly the same. Our discussion is also specific for spin-2 fields.

@@

      Since ${\bf D}^{E\dag}_s{\bf D}^B_s=0$, one can always generate a $B$ mode by applying ${\bf D}^{B}_s$ to a scalar field. It is then possible to show that $E$ and $B$ modes thus defined are orthogonal in the full sky:
      \begin{equation}
        ({\bf D}^E_s\varphi,{\bf D}^B_s\psi)=\int d\nv\,({\bf D}^E_s\varphi)^\dag{\bf D}^B_s\psi=0,
      \end{equation}
      where $\varphi$ and $\psi$ are two arbitrary scalar fields. This can be done by integrating by parts and noting that the celestial sphere has no boundaries. On a cut sky, however, and for $s=2$, this is only true if the fields satisfy Neumann and Dirichlet boundary conditions simultaneously (i.e. vanishing value and first derivative on the boundary of the cut sky region).

      Let $w(\nv)$ be a sky window function defining the sky region to be analyzed (and the weight to be applied in each pixel). The standard pseudo $B$-mode of a spin-2 field ${\bf P}$ is then given by
      \begin{align}\nonumber
        \tilde{B}_{\bf l}&\equiv\int d\nv\,w(\nv)\left(_s{\bf Y}^{B}_{\bf l}(\nv)\right)^\dag{\bf P}\\
                         &=\int d\nv\,w(\nv)({\bf D}^B_sY_{\bf l})^\dag{\bf P}(\nv),
      \end{align}
      Since ${\bf D}^B_sY_{\ell m}$ is a $B$-mode, in the absence of $w$ this expression would correspond to a projection that filters out all the $E$-modes from ${\bf P}$. However, $w(\nv){\bf D}^B_sY_{\ell m}$ is not a $B$-mode, and therefore $\tilde{B}_{\ell m}$ receives contributions from ambiguous $E$ modes (which then propagate into the variance of the pseudo-$C_\ell$ estimator of the power spectrum). The idea behind $B$-mode purification is to move $w$ to the right of ${\bf D}^B_s$, defining the pure $B$ component:
      \begin{equation}
        B^p_{\bf l}=\int d\nv\left({\bf D}^B_s(w\,Y_{\bf l})\right)^\dag\,{\bf P}(\nv).
      \end{equation}
      Since ${\bf D}^B_s(wY_{\bf l})$ is a $B$-mode quantity, $B^p_{\bf l}$ should receive contributions only from $B$-modes.

@@

  \section{Code performance and validation} \label{sec:validation}
    The core of our code-validation suite is based on two science examples. These examples are not meant to be realistic mocks of actual data, however, they do contain the dynamic range and noise-levels of a typical experimental set-up that we expect in the next decade, while providing us with an exact, known true power spectrum to compare against. This section presents these two examples and the results of this validation in terms of accuracy and computational performance. Results will be presented for both curved-sky and flat-sky realizations. Note that, in all cases, we use the true signal power spectra as input to compute the bias from contaminant deprojection (see Eqs. \ref{eq:dp.g.aux} and \ref{eq:dp.f.aux}). This allows us to verify that the code works as expected, and to isolate any residual bias associated with software bugs, rather than an imperfect guess of the true power spectrum.

@@

          \item {\bf Contaminants:} we contaminate the signal maps described above with several different types of residuals. In all cases we add contaminants linearly, to avoid deviating from the model described in Section \ref{ssec:maths.deproj}. The contaminant amplitudes are chosen so that any residual contamination in the relevant range of multipoles will be detected statistically by the validation suite. Maps of these contaminants are shown in the 4 bottom panels of Figure~\ref{fig:lss_mocks_sph} and Figure~\ref{fig:lss_mocks_flat}, and their impact on the angular power spectrum is presented in Fig. \ref{fig:lss_cont_cls} for the curved-sky case.
